---
title: Orchestrating Microbiome Analysis with Bioconductor
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    # These settings below enable word calculation
    filters:
      - at: pre-quarto
        path: _extensions/andrewheiss/wordcount/citeproc.lua
      - at: pre-quarto
        path: _extensions/andrewheiss/wordcount/wordcount.lua
affiliations:
  - id: 1
    name: University of Turku
    department: Department of Computing
    address: Turku, Finland

  - id: 2
    name: University of Turku
    department: Department of Life Technologies
    address: Turku, Finland

  - id: 3
    name: University of Oxford
    department: Department of Biology
    address: Oxford, United Kingdom

  - id: 4
    name: University of Oxford
    department: Queen’s College
    address: Oxford, United Kingdom

  - id: 5
    name: University College Cork
    department: Department of Anatomy and Neuroscience
    address: Cork, Ireland

  - id: 6
    name: University College Cork
    department: APC Microbiome Ireland
    address: Cork, Ireland

  - id: 7
    name: LMU München
    department: Department of Statistics
    address: Munich, Germany

  - id: 8
    name: Cornell University
    department: Department of Population Health Sciences, Weill Cornell Medicine
    address: New York, USA

  - id: 9
    name: Cornell University
    department: Department of Biomedical Informatics & Data Science
    address: New York, USA

  - id: 10
    name: Cornell University
    department: Department of Biostatistics
    address: New York, USA

  - id: 11
    name: Yale University
    department: Department of Biomedical Informatics & Data Science
    address: New Haven, USA

  - id: 12
    name: Yale University
    department: Department of Biostatistics
    address: New Haven, USA

  - id: 13
    name: University of Helsinki
    department: Department of Microbiology
    address: Helsinki, Finland

  - id: 14
    name: Helmholtz Munich
    address: Munich, Germany

  - id: 15
    name: Flatiron Institute
    address: New York, USA

  - id: 16
    name: Institute for Molecular Medicine Finland, FIMM-HiLIFE
    address: Helsinki, Finland

  - id: 17
    name: Danone Research and Innovation
    address: Utrecht, Netherlands

  - id: 18
    name: CUNY Graduate School of Public Health and Health Policy
    department: Department of Epidemiology and Biostatistics
    address: New York, USA

  - id: 19
    name: No affiliation

  - id: 20
    name: Amsterdam University Medical Centres
    department: Department of Psychiatry
    address: Amsterdam, Netherlands

  - id: 21
    name: Amsterdam University Medical Centres
    department: Amsterdam Public Health
    address: Amsterdam, Netherlands

  - id: 22
    name: Cornell University
    department: Division of Gastroenterology and Hepatology, Weill Cornell Medicine
    address: New York, USA

  - id: 23
    name: Cornell University
    department: Department of Statistics and Data Science
    address: New York, USA

author:
  - name: Tuomas Borman
    orcid: 0000-0002-8563-8884
    email: tuomas.v.borman@utu.fi
    affiliations:
      - ref: 1
    attributes:
      corresponding: true

  - name: Giulio Benedetti
    orcid: 0000-0002-8732-7692
    affiliations:
      - ref: 1

  - name: Geraldson Muluh
    orcid: 0000-0002-9721-5192
    affiliations:
      - ref: 1

  - name: Aura Raulo
    orcid: 0000-0003-4860-7840
    affiliations:
      - ref: 3
      - ref: 4

  - name: Benjamin Valderrama
    orcid: 0000-0002-4740-8965
    affiliations:
      - ref: 5
      - ref: 6

  - name: Artur Sannikov
    orcid: 0000-0001-7765-123X
    affiliations:
      - ref: 2

  - name: Stefanie Peschel
    orcid: 0000-0002-7936-7627
    affiliations:
      - ref: 7

  - name: Yihan Liu
    orcid: 0009-0008-0462-7501
    affiliations:
      - ref: 8
      - ref: 9
      - ref: 10
      - ref: 11
      - ref: 12

  - name: Rasmus Hindström
    orcid: 0009-0004-5731-178X
    affiliations:
      - ref: 1

  - name: OMA consortium
    affiliations:
      - ref: 19

  - name: Katariina Pärnänen
    orcid: 0000-0003-2220-1738
    affiliations:
      - ref: 13

  - name: Christian L. Müller
    orcid: 0000-0002-3821-7083
    affiliations:
      - ref: 7
      - ref: 14
      - ref: 15

  - name: Aki S. Havulinna
    orcid: 0000-0002-4787-8959
    affiliations:
      - ref: 1
      - ref: 16

  - name: Sudarshan Shetty
    orcid: 0000-0001-7280-9915
    affiliations:
      - ref: 17

  - name: Marcel Ramos
    orcid: 0000-0002-3242-0582
    affiliations:
      - ref: 18

  - name: Domenick J. Braccia
    orcid: 0000-0002-4606-6254
    affiliations:
      - ref: 19

  - name: Héctor Corrada Bravo
    orcid: 0000-0002-1255-4444
    affiliations:
      - ref: 19

  - name: Felix G.M. Ernst
    orcid: 0000-0001-5064-0928
    affiliations:
      - ref: 19

  - name: Levi Waldron
    orcid: 0000-0003-2725-0694
    affiliations:
      - ref: 18

  - name: Thomaz F. S. Bastiaanssen
    orcid: 0000-0001-6891-734X
    affiliations:
      - ref: 20
      - ref: 21
    attributes:
      equal-contributor: true

  - name: Himel Mallick
    orcid: 0000-0003-4956-2429
    affiliations:
      - ref: 8
      - ref: 22
      - ref: 23
    attributes:
      equal-contributor: true

  - name: Leo Lahti
    orcid: 0000-0001-5537-637X
    email: leo.lahti@utu.fi
    affiliations:
      - ref: 1
    attributes:
      corresponding: true
abstract: |
  The expansion of microbiome research has led to the accumulation of
  interlinked datasets encompassing versatile taxonomic and functional
  assays. The analysis of increasingly large and heterogeneous multi-modal
  microbiome data would benefit from unified approaches supporting the design
  of modular data science workflows through interoperable methods. The
  Bioconductor project recently developed an optimized statistical programming
  framework for multi-assay data integration. Building on this foundation, we
  introduce a community-developed, open-source framework for microbiome data
  science. In contrast to the previous alternatives,
  the methodology is specifically designed to support joint analysis
  of hierarchical, interlinked, and heterogeneous multi-table datasets that
  are increasingly common in modern microbiome research. This
  framework encompasses open data, methods, tutorials, and an active online
  community, thereby supporting standardized and reproducible data
  wrangling, joint analysis, and reporting. We have detailed the functionality
  and usage in the OMA book
  ([https://bioconductor.org/books/release/OMA](https://bioconductor.org/books/release/OMA)),
  which offers guidance for prospective users and contributors.

keywords: [microbiome, Bioconductor, data science, data integration]
bibliography: bibliography.bib
---

# Introduction {#sec-intro}

Modern data science applications critically rely on open-source methods created
by the research community [@barker2022; @pearson2025]. Open
software ecosystems have become central innovation hubs, driving
technological and cultural transformation [@hocquet2024]. The
Bioconductor project is a significant distribution
channel for open data science methods in computational life sciences.
Its vast developer community maintains over 2,300 quality-controlled R
packages for various fields of life science informatics [@gentleman2004].
Additionally, Bioconductor has evolved into a global community that supports
data science collaboration and training [@drnevich2025], playing an increasingly
important role in advancing computational skills that are essential for modern
microbiology education [@timmis2024].

Microbiome research focuses on how microbial communities vary, function and
interact with their environment and hosts [@moreno-indias2021]. The
technologies used in this field generate vast amounts of high-dimensional and
hierarchically structured data, commonly in the form of microbial DNA sequences
detected in a set of samples. Following the rapid expansion of microbiome
research and technological advances, metagenomic sequencing is
complemented by other types of high-throughput measurements. This has
led to the accumulation of multi-modal datasets&mdash;collections
that combine different types of molecular data. These interlinked datasets
encompass diverse taxonomic and functional assays, ranging from microbial
abundances and phylogenetic relationships to functional profiles. These are summarized into tabular data sets whose typical sizes range up to tens of thousands of selected features and samples for downstream statistical analyses.

Microbiome data scientists increasingly face the practical challenges of integrating heterogeneous and high-dimensional observations across multiple data modalities into scalable and reproducible workflows [@ragan-kelley2013;@callahan2016;@holmes2019].
The unique statistical properties and complexity of microbiome data
necessitate specialized approaches [@moreno-indias2021; @jeganathan2021; @willis2019; @bastiaanssen_bugs1_2023; @bastiaanssen_bugs2_2023], yet the diversity of proposed solutions and inconsistent outcomes
[@nearing2022; @pelto2025; @gamboa-tuz2025] can make it
difficult to identify the optimal methods and build interoperable workflows
[@wen2023; @shetty2019]. Accordingly, the field can benefit for standardization and evidence-based best practices [@moreno-indias2021].

Open microbiome data science _frameworks_ have emerged to provide sets of
interoperable methodologies in specific computational environments
[@eren2021; @schloss2009; @bolyen2019; @mcmurdie2012], and a broader ecosystem
of digital resources, engaged user communities and collaboration culture that
extend the capacities far beyond the underlying technical framework.
However, frameworks that are focused solely on microbiome data science fall short on
addressing the need to integrate data across multiple modalities, such as
metagenomics, metabolomics, and host genomics.
Bioconductor integrates the methods of microbiome data science
into a wider statistical programming framework, where
interoperable methods for single-cell omics [@amezquita2020], spatial transcriptomics [@crowell2025],
metabolomics and proteomics [@anwar2025], and other
fields are concurrently developed.

This approach can introduce some challenges, such as dependency management, package compatibility, and differences in package stability. These issues are addressed through coordinated mechanisms that help to ensure the robustness of the software ecosystem. First, submitted packages undergo peer review, ensuring relevance and adherence to standards such as unit tests, reproducible documentation, and working examples. Secondly, Bioconductor follows a synchronized, semiannual release cycle, during which packages are frozen and mutually compatible. Thirdly, all packages are continuously tested within Bioconductor's automated build system, which ensures that breaking changes are rapidly detected before release. Each package has an active maintainer who is notified of failures and responsible for resolving them in a timely manner. Finally, community guidelines and continuous integration practices, together with shared data structures such as (Tree)SummarizedExperiment, further improve interoperability between packages.

Here, we introduce a stable and optimized data science ecosystem supporting the
integration and analysis of multi-modal datasets in microbiome research.
It provides harmonized data import and representation, and fast,
robust methods for transforming abundance tables into biological insights. These
methods have been extensively tested and distributed through mature R/Bioconductor
packages
([https://bioconductor.org/packages](https://bioconductor.org/packages)),
supporting the community-driven development and integration
within the broader Bioconductor ecosystem. To promote evidence-based best
practices, we introduce the online book _Orchestrating Microbiome Analysis with
Bioconductor_ (OMA), which represents a comprehensive compilation of established
approaches and emerging trends in microbiome data science using the
R/Bioconductor ecosystem.

# Data infrastructure {#sec-infra}

Data scientists routinely deal with interlinked assays, hierarchical data
(*e.g.*, ontologies, trees), relational data (*e.g.*, networks), and supporting
side information. This side information can include
sample attributes such as treatment group, collection site, or time point. The
ability to integrate heterogeneous data elements is essential in biological
research, where the individual components of a system can seldom be studied in
isolation. However, this growing complexity introduces additional overhead, as
researchers must track samples and features across multiple tables and manage an
increasing number of data elements and non-trivial links between them. Moreover,
input data formats vary across methods, leading to technical
complications and time lost on data wrangling. Bioconductor addresses these
challenges through standardized data structures that are extended by an
ecosystem of quality-controlled, interoperable methods. Adopting the framework will thus allow the analyst to avoid unnecessary data wrangling and invest more time in core analytical tasks.

Bioconductor data structures support statistical analysis of complex
data combinations [@huber2015] and are widely adopted in omics research (@fig-bioc_packages).
They have been implemented through specialized
object-oriented classes that integrate multiple data elements into
a single structure known as a *data container*. The SummarizedExperiment (SE) class [@huber2015] is
the primary data container underlying the Bioconductor ecosystem
(Figure \ref{fig:datacontainers}a and @tbl-elements).
Its position among the top 1\% most-downloaded
packages in Bioconductor reflects its widespread recognition and
increasing adoption among developers. It provides a
standardized solution to store tabular data by linking numeric abundance
matrices with rich side information on the features/rows  (*e.g.*, taxonomy),
and samples/columns (*e.g.*, origin, collection time, host health).

The vast adoption and interoperability of SE-based data structures
enables the transfer of methods across different fields of omics research,
thereby promoting the development of modular and scalable data science methods
and yielding improved overall quality control and user support [@drnevich2025].
The SE framework has been extended to similar data integration tasks, notably in single-cell and spatial omics through the SingleCellExperiment (SCE) [@amezquita2020] and SpatialExperiment [@righelli2022] classes.

```{r}
#| label: bioc_packages
#| eval: false
#| echo: false
#| message: false

if (!require("BiocManager")) {
    install.packages("BiocManager")
    library("BiocManager")
}
pkgs <- c("BiocPkgTools", "lubridate", "tidyverse")
temp <- sapply(pkgs, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
        install(pkg)
        library(pkg, character.only = TRUE)
    }
})

# Get packages and info based on their BiocViews
pkgs <- biocPkgList()
# Get how many times these packages are loaded
df <- biocDownloadStats()

# Take only packages that depend on SE
pkgs[["SE"]] <- sapply(pkgs[["Depends"]], function(x) any(grepl("summarizedexperiment", x, ignore.case = TRUE)))
pkgs <- pkgs[pkgs[["SE"]], ]

# Extract the package topic based on biocView
fields <- sapply(pkgs[["biocViews"]], function(x) {
    field <- "Other"
    if (any(grepl("genomic|genetic|gene|genom|DNA", x, ignore.case = TRUE))) field <- "Genomics"
    if (any(grepl("proteomics|protein", x, ignore.case = TRUE))) field <- "Proteomics"
    if (any(grepl("metabolomics|metabolome|metabolite|Lipidom|massspectro", x, ignore.case = TRUE))) field <- "Metabolomics"
    if (any(grepl("transcripto|RNA-seq|RNA", x, ignore.case = TRUE))) field <- "Transcriptomics"
    if (any(grepl("immuno", x, ignore.case = TRUE))) field <- "Immunology"
    if (any(grepl("cytom", x, ignore.case = TRUE))) field <- "Cytometrics"
    if (any(grepl("microarray|chip", x, ignore.case = TRUE))) field <- "Microarray"
    if (any(grepl("single-cell|singlecell", x, ignore.case = TRUE))) field <- "Single-cell"
    if (any(grepl("metagenom|microbiome|16S|microbiota|amplicon|shotgun|microb|metatranscript|metametabolo|metaproteo", x, ignore.case = TRUE))) field <- "Microbiome"
    return(field)
})
pkgs[["Field"]] <- fields

# Subset downloads table and add field to it
df <- df[df[["Package"]] %in% pkgs[["Package"]], ]
df[["Field"]] <- pkgs[match(df[["Package"]], pkgs[["Package"]]), "Field"][[1]]

# Get information on when the package became available in Bioconductor
dates <- df[["Date"]] |> unique()
dates <- dates[dates < floor_date(Sys.time(), "month")]
available <- lapply(dates, function(date) {
    temp <- df[df[["Date"]] == date, ]
    temp <- temp[match(unique(df[["Package"]]), temp[["Package"]]), ]
    temp <- temp[["Nb_of_distinct_IPs"]] > 0 | temp[["Nb_of_downloads"]] > 0
    temp[is.na(temp)] <- FALSE
    as.numeric(temp)
})
available <- do.call(cbind, available) |> as.data.frame()
colnames(available) <- dates
rownames(available) <- unique(df[["Package"]])

# Create a table showing the number of packages by field through time
ind <- pkgs[match(unique(df[["Package"]]), pkgs[["Package"]]), "Field"][[1]]
pkgs_date <- rowsum(available, group = ind)
pkgs_date <- pkgs_date |>
    rownames_to_column("Field") |>
    pivot_longer(
        cols = -Field,
        names_to = "Date",
        values_to = "N"
    )
pkgs_date[["Date"]] <- as.Date(pkgs_date[["Date"]])

# Save results
pkgs_date |>
    write.csv(file = file.path("data", "bioc_packages_stats.csv"), row.names = FALSE)
```

```{r}
#| label: fig-bioc_packages
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 6
#| fig-cap: "**Adoption of the SummarizedExperiment (SE) data container among
#|   Bioconductor developers.** The growth in the number of Bioconductor
#|   packages supporting SE over time is shown, categorized by their respective
#|   fields. Packages are counted by the date they were added to Bioconductor.
#|   Some packages added the SE support after their original release."

if (!require("BiocManager")) {
    install.packages("BiocManager")
    library("BiocManager")
}
pkgs <- c("tidyverse", "see")
temp <- sapply(pkgs, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
        install(pkg)
        library(pkg, character.only = TRUE)
    }
})

# Import data
pkgs_date <- read.csv(file.path("data", "bioc_packages_stats.csv"))
pkgs_date[["Date"]] <- pkgs_date[["Date"]] |> as.Date()

# Create the plot
p <- ggplot(pkgs_date, aes(x = Date, y = N, fill = Field)) +
    geom_area() +
    theme_classic() +
    scale_fill_okabeito() +
    labs(x = "Year", y = "Number of packages") +
    scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
    scale_x_date(expand = c(0, 0), limits = c(as.Date("2015-01-01"), max(pkgs_date[["Date"]])))

p
```

**Hierarchical data structures.** Microbiome data scientists frequently need to
integrate information on feature phylogenies and sample hierarchies
to accurately reflect fine-scale variability in microbiome data.
While maintaining interoperability with
the broader Bioconductor ecosystem, the TreeSummarizedExperiment
(TreeSE) class [@huang2021] extends the SE and SCE data structures to
hierarchical datasets by incorporating both feature and sample organizations
as tree structures (Figure \ref{fig:datacontainers}b and @tbl-elements).
Beyond the core components of microbiome datasets, including taxa abundances,
sample metadata, phylogenetic trees, and reference sequences, TreeSE provides a
flexible foundation for multi-modal data integration, supporting the incorporation of functional predictions, resistome, metabolome, and other
molecular profiles, as well as alternative feature representations.
It can also store results from common analytical procedures, such as statistical transformations or agglomeration. By inheriting full compatibility with SE and SCE, TreeSE enables the development of modular, interoperable workflows across the Bioconductor ecosystem.


::: {.figure}
\begin{center}
\includegraphics[height=\textheight]{figures/data_containers.png}
\end{center}
\captionof{figure}{\textbf{Integrative Bioconductor containers for microbiome data.}
\textbf{(a) SummarizedExperiment (SE)} \cite{huber2015} is a generic container
for single-omics experiments. It accommodates \textit{assays} that describe the
feature abundances in each sample (\textit{e.g.}, taxonomic units, resistance
genes, or predicted functions), and tabular side information on each sample
(\textit{colData}) and feature (\textit{rowData}), \textit{e.g.}, the taxonomic
classification.
\textbf{(b) TreeSummarizedExperiment (TreeSE)} \cite{huang2021} supports
hierarchical, multi-table microbiome analysis. In addition to the SE elements,
TreeSE incorporates tree information on feature hierarchies, such as
phylogenetic relations between the taxonomic units (\textit{rowTree}) or host
species (\textit{colTree}), and results of dimensionality reduction
(\textit{reducedDims}). TreeSE can thus link multiple abundance tables
and their transformed versions (\textit{e.g.}, statistical transformation,
taxonomic agglomeration, or dimensionality reduction) with tabular and
hierarchical side information on the features (rows) and samples (columns).
Additional slots for reference sequences and experiment metadata are available.
\textbf{(c) MultiAssayExperiment (MAE)} \cite{ramos2017} facilitates the
integration of interlinked datasets representing different omics modalities
linked by potentially non-trivial sample mappings. Figure adapted from \cite{ramos2017}.}
\label{fig:datacontainers}
:::

**Multi-assay data structures.** Contemporary microbiome research frequently
involves data across multiple measurement modalities for enhanced functional
and mechanistic insights. This brings the need to find an appropriate
representation for each data type and map the links between them. Whereas the
TreeSE data container is broadly applicable to different modalities such as
taxonomic, resistome, transcriptome, proteome, and metabolome profiling data,
dedicated data containers have been developed for certain modalities by the
Bioconductor community to address their specific characteristics; examples
include single-cell sequencing [@amezquita2020] and host genomics
[@lawrence2013]. Linking data across modalities presents an additional technical
challenge, as multi-omics datasets are often sparse so that only a subset of the
samples may be shared across multiple modalities [@sherwani2025]. In other
cases, a single sample in one modality may link to several samples in
another modality [@husso2023] (*e.g.*, due to technical or spatial replication).
These scenarios require the ability to integrate multiple
data modalities into a single object and create flexible sample mappings across them.
The MultiAssayExperiment (MAE) data container [@ramos2017] addresses these
challenges by introducing a formal mapping layer (sampleMap) that links
biological samples to their corresponding measurements in each modality
(Figure \ref{fig:datacontainers}c and @tbl-elements).
This design supports one-to-one, one-to-many, and
partially overlapping sample relationships across experiments. Importantly, this
mapping is also accessible for downstream operations. For instance, when
subsetting a MAE object
(*e.g.*, selecting samples with both microbiome and metabolome data), the
framework automatically resolves the intersection of available samples and
ensures consistent alignment across all modalities. Such structured handling of
sample relationships is particularly important for integrative methods, as it
reduces the risk of sample mismatches, a common source of irreproducibility in
ad hoc multi-table analyses. To summarize, flexible mapping of samples
across several data objects is a key for efficient data integration, method sharing,
and interoperability.

**Scalability.** Enhanced scalability is one of the advantages of our framework, and essential with increasingly large and multi-modal data sets. This can be achieved by parallelization, out-of-memory analyses leveraging the built-in support for sparse and delayed matrices, more efficient implementations through code optimization, executing costly computations in C++, and methodological advancements. In many cases, the support can be readily inherited from the underlying Bioconductor infrastructure, rather than newly implemented [@lun2018beachmat]. For instance, the (Tree)SE and MAE containers are agnostic to assay representation and natively support sparse (dgCMatrix, SparseArray) and file-backed (HDF5Array, DelayedArray) assays [@huber2015]. Many methods readily support parallelization; BiocParallel can support operations that decompose into independent blocks by samples, features, or other data elements. Non-decomposable operations, such as distance matrices and ordinations, must be realized in memory or parallelized by a dedicated algorithm: accelerated versions of common microbiome methods, including Faith phylogenetic alpha diversity [@armstrong2021] and striped UniFrac beta diversity [@mcdonald2018], have been incorporated in the (Tree)SE-based framework through the mia package.


| Slot          | Content                                      | SE | TreeSE | MAE$^a$ |
|---------------|----------------------------------------------|:--:|:------:|:-------:|
| **Tables**    |                                              |    |        |         |
| assays        | tables of features (rows) and samples (cols) | X  | X      | X       |
| altExps$^b$   | experiments with variable number of features |    | X      | X       |
| reducedDims   | results of dimensionality reduction          |    | X      | X       |
| **Rows**      |                                              |    |        |         |
| rowData       | side information on features                 | X  | X      | X       |
| rowPairs      | linkage between pairs of associated features |    | X      | X       |
| rowTrees      | hierarchical organization of features        |    | X      | X       |
| referenceSeq  | reference sequences of features              |    | X      | X       |
| **Columns**   |                                              |    |        |         |
| colData       | side information on samples                  | X  | X      | X       |
| colTrees      | hierarchical organization of samples         |    | X      | X       |
| sampleMap$^c$ | flexible sample mapping across experiments   |    |        | X       |
| **Other**     |                                              |    |        |         |
| metadata      | additional information on the experiment     | X  | X      | X       |

: Key elements of the Bioconductor data containers for microbiome analysis.
{#tbl-elements tbl-colwidths="[18,62,3,9,8]"}

$^a$While MAE does not contain conventional slots (with the exception
of colData), its experiments can include any number of data containers that
provide those slots. $^b$altExp samples must match assay samples in a 1-to-1
mapping. $^c$The sampleMap supports diverse mappings between samples from
different experiments (1-to-1, 1-to-many, many-to-1 and many-to-many). SE,
SummarizedExperiment; TreeSE, TreeSummarizedExperiment; MAE, MultiAssayExperiment.

**Data resources.** Several open microbiome data resources are supported by this framework, enabling direct data import into the aforementioned data containers for further downstream analysis within R/Bioconductor (@tbl-dataresources). Examples include curatedMetagenomicData [@pasolli2017], HoloFood [@rogers2024],
microbiomeDataSets (Bioconductor package), and the EBI/MGnify database
[@gurbich2023]. Bioconductor packages include additional built-in demonstration datasets. Collectively, these resources encompass taxonomic and functional profiles from thousands of published microbiome studies and tens of thousands samples that can
be accessed with the available tools for research and educational purposes.

| Data resource / Package     | References                 | # Studies  | # Samples |
|-----------------------------|:--------------------------:|-----------:|----------:|
| MGnifyR$^a$                 | [@gurbich2023]             | 5,616      | 429,108   |
| Metalog$^b$                 | [@kuhn2026]                | 1,074      | 157,483   |
| curatedMetagenomicData$^a$  | [@pasolli2017]             | 93         | 22,588    |
| miaData$^c$                 |                            | 9          | 20,404    |
| microbiomeDataSets$^a$      |                            | 6          | 19,100    |
| HMP2data$^a$                |                            | 3          | 11,493    |
| HoloFoodR$^{a,d}$           | [@rogers2024; @borman2025] | 9          | 9,990     |
| MicrobiomeBenchmarkData$^a$ | [@gamboa-tuz2025]          | 6          | 1,125     |
: Open microbiome data resources (last accessed on 2026-07-05).
{#tbl-dataresources tbl-colwidths="[40,20,20,20]"}

$^a$Bioconductor package with the corresponding name.
$^b$Data retrieval to TreeSE format instructed in OMA online book.
$^c$Zenodo mia community.
$^d$HoloFood microbiome datasets are deposited in the MGnify database.
HMP, Human Microbiome Project.

# Alternative frameworks {#sec-data_frameworks}

Several alternative frameworks to microbiome analysis are available. Within Bioconductor, phyloseq [@mcmurdie2012] provides widely adopted tools with a primary focus on 16S amplicon data analysis. (Tree)SE accommodates all core components of phyloseq objects, including abundance tables, taxonomic annotations, phylogenetic trees, reference sequences, and sample metadata. The advantages of newer (Tree)SE container [@huber2015] include many additional data elements, a better integration with the overall Bioconductor ecosystem, and enhanced support for multi-assay analyses (see [Data infrastructure](#sec-infra)). Thus, some originally phyloseq-based tools, such as microbiome R/Bioconductor package [@shetty2019], have been recently extended with TreeSE support.

Related frameworks beyond R/Bioconductor include command-line toolkits supported by active user and developer communities. Mothur [@schloss2009] is primarily designed for 16S amplicon analysis and less well suited for integrative metagenomics and multi-omics. QIIME 2 [@bolyen2019] is a Python-based, AI-ready microbiome multi-omics data science platform. It primarily targets amplicon and metagenomic analyses, and incorporates methods from scikit-bio Python library [@aton2026] which supports multi-omic integration tasks. A particular advantage of these frameworks is their direct connection with the strong machine learning and AI ecosystem in Python, while the interoperability with independently developed omics tools may remain more limited. Anvi'o [@eren2021] is a complementary, community-driven analysis and visualization platform for microbial 'omics with a distinct focus on genome-resolved metagenomics and detailed sequence-level analyses. These frameworks emphasize standardized, modular approaches tailored to microbiome analysis and provide validated and user-friendly solutions, supporting efficient end-to-end workflows for microbiome research. Bioconductor emphasizes broad interoperability and shared statistical programming infrastructure with other areas of omics research, and primarily serves the R community. To facilitate migration from alternative frameworks, the OMA book provides conversion tables and utility functions to convert between TreeSE and many external data formats.

Quantitative comparisons indicate enhanced scalability over commonly used alternatives (@fig-benchmarking and Supplementary Figure 1). We executed routine operations on increasingly large subsets of the Metalog database [@kuhn2026], and measured the execution time and memory consumption with TreeSE, phyloseq, speedyseq, and QIIME 2. Overall, (Tree)SE scaled most efficiently in execution time and memory allocation with increasing feature and sample sizes. It is applicable to abundance matrices of up to 100,000 samples and 10,000 features on a standard laptop, except for the heaviest operations such as phylogenetic beta diversity (UniFrac) calculation, where the general break point is reached around 10,000 samples with all methods. At a feature set size of 10,000 the break point is additionally reached with the PhILR transformation and melting operation. At smaller sample sizes phyloseq or its accelerated extension speedyseq are occasionally faster but these differences are moderate (few seconds on a standard laptop). QIIME 2 is generally slower on lower sample sizes but scales well with increasing sample sizes. Despite the advantages, optimizing time and memory consumption represents a key area for further development.




```{r}
#| label: fig-benchmarking
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "**Execution time and allocated memory for routine microbiome data
#|   operations.** TreeSummarizedExperiment, QIIME 2 and phyloseq&mdash;with its
#|   extension speedyseq&mdash;are three popular frameworks for microbiome
#|   analysis. We benchmarked them for five routine operations applied to random
#|   subsets of features (rows) and samples (columns) from the Metalog database
#|   [@kuhn2026]. Both execution time (s) and allocated memory (MB) were
#|   measured as the mean value over ten replicates with varying subsets of the
#|   same dimension. Errorbars show the 95% confidence interval of the
#|   performance estimate. The benchmarking was conducted in R using the bench
#|   library with 4 CPU cores and 16 GB RAM. QIIME 2 lacks allocated memory
#|   estimates because memory cannot be monitored for external processes. The
#|   source code for the benchmarking analysis is openly available in the Zenodo
#|   mia community."
#| fig-width: 12
#| fig-height: 14

# Adapted from https://github.com/microbiome/benchmarking/tree/main/

# Import libraries
if (!require("BiocManager")) {
    install.packages("BiocManager")
    library("BiocManager")
}

pkgs <- c("ggplot2", "patchwork", "tools")

temp <- sapply(pkgs, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
        install(pkg)
        library(pkg, character.only = TRUE)
    }
})

# Define class names
classes <- c(
    tse = "TreeSE",
    qiime = "QIIME 2",
    pseq = "phyloseq",
    spseq = "speedyseq"
)

# Define method names
methods <- c(
    trans = "PhILR transformation",
    agg = "Family agglomeration",
    alpha = "Faith diversity",
    beta = "UniFrac dissimilarity",
    melt = "Melting"
)

# Import results
df <- file.path("data", "benchmark_results.csv") |>
    read.table(sep = ",", header = TRUE)

# Convert object and method vars to factors
df$object <- factor(df$object, levels = names(classes))
df$method <- factor(df$method, levels = names(methods))

# Specify plot layouts
scientific_10 <- function(y) {
    sapply(y, function(z) {
        if (is.character(z)) {
            z <- as.numeric(z)
        }
        if (is.na(z)) {
            NA
        } else if (z %in% c(1, 10)) {
            as.character(z)
        } else {
            paste0("10^", log10(z))
        }
    })
}

label_scientific <- function(x) parse(text = scientific_10(x))

col.breaks <- unique(df$cols[log10(df$cols) %% 1 == 0])

text_col <- get_theme()$axis.text$colour

cus_theme <- theme(
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 15),
    legend.key.size = unit(1.2, "cm"),
    axis.title = element_text(size = 15),
    axis.text.y.right = element_blank(),
    axis.ticks.y.right = element_blank(),
    axis.text = element_text(size = 12),
    strip.text.x = element_text(size = 15),
    strip.text.y = element_text(colour = text_col, size = 12, angle = 0),
    strip.background = element_blank(),
    plot.tag = element_text(size = 16, face = "bold")
)

# Visualise benchmarking results: time
plot_bench <- function(df, bench.var, error.var = "SE") {
    var_name <- toTitleCase(bench.var)
    df$Mean <- df[[var_name]]
    df$Err <- df[[paste0(var_name, error.var)]]

    axis_title <- switch(bench.var,
        time = "Execution time (s)",
        memory = "Allocated memory (MB)"
    )

    y.breaks <- switch(bench.var,
        time = 10^seq(-1, 3),
        memory = 10^seq(0, 4)
    )

    y.lims <- range(y.breaks)
    y.lims <- c(min(y.lims[1], min(df$Mean)), max(y.lims[2], max(df$Mean)))

    row.breaks <- unique(df$rows[log10(df$rows) %% 1 == 0])
    df$rows <- scientific_10(df$rows)

    p <- ggplot(df, aes(x = cols, y = Mean, colour = object)) +
        geom_errorbar(
            aes(ymin = Mean - Err, ymax = Mean + Err),
            width = 0, show.legend = TRUE
        ) +
        geom_line(show.legend = TRUE) +
        geom_point(show.legend = TRUE) +
        scale_x_log10(breaks = col.breaks, limits = range(df$cols), labels = label_scientific) +
        scale_y_log10(
            labels = label_scientific,
            breaks = y.breaks,
            limits = y.lims,
            sec.axis = sec_axis(~., name = "# Features")
        ) +
        scale_colour_manual(
            labels = classes,
            values = c("firebrick", "mediumblue", "black", "darkgrey"),
            drop = FALSE
        )

    grid_by <- ". ~ method"
    grid_lab <- list(method = methods)

    if (length(row.breaks) > 1L) {
        grid_by <- sub(".", "rows", grid_by, fixed = TRUE)
        grid_lab[["rows"]] <- label_parsed
    }

    p <- p +
        facet_grid(
            as.formula(grid_by),
            labeller = do.call(labeller, grid_lab)
        )

    p <- p +
        labs(x = "# Samples", y = axis_title, colour = "Object") +
        theme_bw() +
        cus_theme

    return(p)
}

# Visualise benchmarking results
p1 <- plot_bench(df, "time", "CI")
p2 <- plot_bench(subset(df, Memory <= 1e3 * 16), "memory", "CI")

# Combine results
p <- (p1 / p2) +
    plot_layout(guides = "collect") &
    plot_annotation(tag_levels = "a") &
    cus_theme

p
```

# Data preparation {#sec-process}

While several tools for microbiome data analysis are available [@jeganathan2021; @willis2019; @marcoszambrano2023], their implementations typically differ in terms of expected input and output formats. Common data standards promote modular workflows, speeding up methods development, benchmarking, and replication. Building upon these principles, we describe a set of interoperable software libraries that support microbiome data integration and analysis based on the two integrative data containers, TreeSE and MAE (Figure \ref{fig:workflow}).

This framework specifically supports the downstream analysis of high-throughput sequencing data from microbiome experiments, after summarizing the original measurements into abundance tables (*e.g.*, taxonomic or functional profiles). Taxonomic abundance tables are one of the most commonly encountered data types in contemporary microbiome studies. Common measurement platforms include amplicon-based marker gene studies (*e.g.*, 16S rRNA) and metagenomic sequencing, processed with methods such as MetaPhlAn [@blanco2023] or Greengenes2 [@mcDonald2024Greengenes2]. Alternative profiling techniques are available, including phylogenetic microarrays [@rajilicstojanovic2009] and high-density optical mapping [@abakumov2024]. Tabular abundance data can also arise from resistome profiling, mapping of other genomic elements, or from functional measurements. The framework is agnostic to the source of abundance data but many of our methodological choices have been guided by data tables that are typical for microbiome profiling studies.


**Data import.** While the TreeSE data container can be populated with microbiome data from common tabular data types, non-standard file formats may
require additional data wrangling. To streamline data
import, we provide importers for a variety of standard microbiome file
formats, including BIOM [@mcdonald2012], MetaPhlAn/HUMAnN (versions 2-4) [@beghini2021], Mothur [@schloss2009], QIIME 2 [@bolyen2019], and the TAXonomic Profile Aggregation and standardization format (taxpasta) [@beber2023]. External to Bioconductor, the qiime2R package provides additional functionality to create TreeSE objects directly from QIIME 2 artifacts.

These importers retain provenance of the metadata and the conversion workflow, bridging the gap between the original data processing, other tools, and downstream statistical analysis in R. This allows researchers to focus on interpreting biological patterns rather than wrangling file formats. Moreover, converters between alternative data structures in R are available supporting seamless conversions between TreeSE and BIOM, DADA2 [@callahan2016], and phyloseq (the mia::convert\* functions). Imported datasets from different omics modalities can then be interlinked within the MAE data container with its flexible sample mappings as described above. This integrated data object can then be analyzed with dedicated downstream methods.

Similarly, microbiome data stored in TreeSE objects can be
exported to external services such as MicrobiomeDB [@oliveira2018] and to other languages (*e.g.*, Julia). These converters facilitate widespread access to
microbiome analysis methods.

**Quality control and exploration.** After importing the data into R,
the recommended first step involves basic data exploration and quality
control [@zuur2010]. In addition to methods for identifying
contaminant sequences and calculating summary statistics, the
framework includes a versatile set of custom visualization methods to
assess data variability, outliers, homogeneity of variance and other
aspects that may need to be considered in downstream data
analysis. The mia and miaViz packages provide
convenient access to many operations for the integrative
data containers regarding common tasks in microbiome
data exploration and visualization (@fig-downstream_analyses).

**Filtering and agglomeration.** Microbiome datasets can contain tens of
thousands of unique features depending on the sequencing resolution, used
reference databases, and the scope of analysis. Moreover, the integration of
multiple omics modalities adds another level of challenges. To summarize data
into biologically meaningful groups or to reduce multiple comparisons the data
is often subsetted, filtered or agglomerated to higher-level taxonomic or
functional categories or stratified by sample groups. The package ecosystem,
and the mia package in particular, feature various tools for such tasks.
For instance, the prevalence-based methods can be
helpful to remove rare features that are likely sequencing artifacts
or present otherwise limited value for the analysis. The mechanism of
alternative experiments (altExp) allows agglomeration
by taxonomic levels or other features such as
pathogenicity or functional potential (*e.g.*, butyrate production
[@kullberg2024], gut-brain modules [@vallescolomer2019], and
host-associated signatures as in the BugSigDB [@geistlinger2024]). The agglomerated data can be
incorporated as a new alternative experiment within the original data object.
The automated tracking of feature and sample links ensures that
changes are propagated across all relevant elements of the data object,
such as transformed abundance tables, phylogenetic trees, or interlinked
omics modalities. This helps to avoid redundant processing and enables
efficient management of interlinked datasets.

**Transformation.** Statistical transformations commonly used in microbiome
research include relative abundance [@gloor2017],
phylogenetic ILR (PhILR) [@silverman2017], centered log-ratio (CLR),
and robust CLR (rCLR) [@martino2019]. The latter three aim to mitigate
compositionality bias. A related concept that controls for
unequal sampling depth is data rarefaction, i.e. repeatedly subsampling data to equal read depth and averaging the computed metrics to mitigate information loss [@schloss2024b] that could arise from subsampling (*rarefying*) data to equal read depth without such iteration [@mcmurdie2014; @schloss2024; @willis2019b]. Our framework supports these and a variety of other transformations commonly applied to microbiome data.

**Data enrichment.** Incorporating additional data, such as transformed
feature abundance tables, is a frequent operation in applied data analysis, and it often takes place iteratively. The integrative data containers
support gradual and organized accumulation of interlinked data elements
including alpha and beta diversity, prevalence calculations, transformations, agglomerated data, reduced dimensions, and other derived data on samples and features. We generally recommend initializing the data objects by pre-calculating such measures already at the beginning of the workflow, where they can then be gradually enriched as the project advances. By systematically extending the upstream data preparation step, instead of ad-hoc calculations scattered amidst the workflow, one can achieve a well-organized and transparent provenance of the integrated data object.


::: {.figure}
\begin{center}
\includegraphics[height=\textheight]{figures/workflow.png}
\end{center}
\captionof{figure}{\textbf{Multi-modal data integration workflow.} Following
data generation and preprocessing (top two panels), the integrative framework
covers five steps of microbiome data integration and analysis: 1) Data is
combined into a TreeSummarizedExperiment (TreeSE) data container,
storing information on the samples (tube icons) and their features
(\textit{e.g.}, taxonomic, functional, or resistome profiles; colored circles).
TreeSE objects can accommodate multiple assay tables as well as additional
hierarchical structures. 2) The TreeSE data object can then be processed to clean
the data. This involves filtering out low-quality samples and features,
transforming assays for example by normalizing feature-wise to enhance the
comparability of sample-specific abundances, or agglomerating features
to specific taxonomic levels or functional categories. 3) The cleaned data can
then be analyzed using statistical models to uncover relations among samples
(\textit{e.g.}, with ordination methods), analyze drivers of variation in community diversity
and composition, or test for differential abundance (\textit{e.g.} between
sampling time or location). 4) The framework supports the integration of
various types of omics modalities through the MultiAssayExperiment and
omic-specific data containers provided by Bioconductor.
For example, the taxonomic and metabolomic profiles of samples can be summarized
and compared in a joint analysis among features. 5) Finally, Quarto notebooks
aid reproducible analysis and reporting.}
\label{fig:workflow}
:::

# Downstream data analysis {#sec-analysis}

Our framework has been designed to improve the management of integrative multi-omic workflows. It facilitates joint operations on interlinked data elements, helping to shift the focus towards multi-modal analyses and
methods development while also supporting standard data analyses in
microbiome research.

Microbiome data analysis often relies on specialized statistical approaches,
since data sparsity, zero-inflation, compositionality, and heteroscedasticity
pose challenges for standard methods [@gloor2017; @jeganathan2021].
Furthermore, microbiome data is hierarchically structured across taxonomic, ecological, temporal, and spatial levels [@moreno-indias2021; @sherwani2025; @ruuskanen2021]. Alternative processing pipelines introduce different biases and generate varying data formats; for example, the same sample can yield differing taxonomic profiles depending on the computational method used [@quince2017]. These and other unique characteristics have wide-reaching implications for microbiome analysis, necessitating dedicated analysis methods [@moreno-indias2021; @jeganathan2021]. The shared data containers support community-driven development and reproducible benchmarking.

The following sections provide an overview of some of the currently available tools, encompassing standard methods for taxonomic, metagenomic, and functional abundance tables as well as integrative analyses between modalities.

**Community indicators.** Various general indicators are available
to quantify the microbial community state or other sample
properties. Community diversity is a common measure in
microbial ecology [@bastiaanssen_bugs1_2023]
(@fig-downstream_analyses). Alpha diversity metrics generally quantify
the observed or estimated number of distinct taxa (_richness_) and
their distribution (_evenness_). Many diversity indices (*e.g.*,
Shannon index), combine these two aspects. Phylogenetic diversity
metrics additionally account the breadth of phylogenetic relatedness among
community members, but are more computationally demanding than
abundance-based metrics. A recent implementation, the Stacked Faith's
Phylogenetic Diversity, improves computational efficiency by 1&ndash;2
orders of magnitude [@armstrong2021] by an optimized single-pass algorithm and utilizing data sparsity. The framework supports analyses with repeated rarefaction through the mia package [@schloss2024b]. Other
global indices of community state include measures of dominance,
rarity, skewness, kurtosis, modality, library size (number of reads),
or absolute abundance quantification. Sample information in the TreeSE object
(colData) can be enriched with these derived quantities. By default,
the mia::addAlpha method returns a set of complementary alpha
diversity measures [@cassol2025]. Alpha diversity indices have been frequently
applied to measure also resistome diversity [@lee2023; @salehi2025], and potentially other modalities.

**Community similarity.** This is commonly evaluated with beta diversity
measures, which can be used for quantitative
comparisons of multivariate community composition. Our
framework supports common dissimilarity metrics in microbial ecology,
including Bray-Curtis, Jaccard, Aitchison
[@bastiaanssen_bugs1_2023], and the accelerated phylogeny-aware (striped) UniFrac [@mcdonald2018]. Community dissimilarity is commonly visualized with unsupervised ordination that projects data onto a lower-dimensional space. Among unsupervised techniques, the most commonly used ones
include Principal Component Analysis (PCA) and Principal Coordinate Analysis
(PCoA) (@fig-downstream_analyses). Statistical tests
(*e.g.*, PERMANOVA) and supervised
ordination methods, such as
distance-based Redundancy Analysis (dbRDA), can complement the
unsupervised analyses, and help uncover and quantify covariation
between community composition and sample-specific covariates. This type of analysis can benefit
from compositionally robust measures (*e.g.*, robust
Aitchison distance [@martino2019]), or phylogeny-aware dissimilarity
measures specifically developed for microbiome research (*e.g.*, UniFrac). Sample dissimilarity provides the basis for many
specialized microbiome techniques, including microbial community typing
(clustering) with Dirichlet multinomial mixtures [@holmes2012]
and community state types [@digiulio2015], or factorization into
community signatures [@frioux2023]. These and other multivariate
methods are provided by mia, miaViz, and additional single-cell packages
(scater, scuttle [@amezquita2020]). Many omics modalities can benefit from these and other standardized methods.

**Association analyses.** Differential Abundance Analysis (DAA)
and Differential Prevalence Analysis (DPA) methods identify indicator microbial
features whose abundances (DAA) or prevalences (DPA) vary across study groups
or continuous gradients, for example in terms of spatial or temporal dissimilarity between
samples (@fig-downstream_analyses). Many widely used DAA methods support the (Tree)SE framework, including for instance ANCOM-BC2 [@lin2024], LimROTS [@anwar2025], MaAsLin3 [@nickols2026], and radEmu [@clausen2026],
and benchdamic benchmarking pipeline [@calgaro2022].
More recently, DPA methods have seen increasing adoption in microbiome
research. They ignore abundance data, and focus on comparing feature
prevalence i.e. presence/absence profiles. Recent implementations
include MaAsLin3 [@nickols2026] and the probabilistic DiPPER
[@pelto2026].
Leveraging the interoperability of the framework, several of these methods have also found applications with other omics modalities, including transcriptomics
[@love2014], metabolomics [@nickols2026], and proteomics [@anwar2025].
Standard DAA typically focuses on
individual features, ignoring interactions or other covariation among feature
abundances. Thus, we support covariation analyses. Recent studies have indicated that elementary methods, such as log-transformed Total Sum Scaled (TSS) counts coupled with a linear model, or presence/absence data with logistic regression, often yield the most consistent results [@pelto2025; @gamboa-tuz2025].
While agglomerating collinear
features as a preprocessing step can reduce multiple testing, increase
statistical power, and improve interpretability, it is also prone to subjective
parameterization and limited generalizability. Features can be grouped
based on clustering results from
abundance profiles or on external information drawn from phylogenetic
relations, taxonomic classifications or functional properties
(*e.g.*, aggregating based on metabolite production [@kullberg2024] or disease associations [@geistlinger2024]). Beyond these, recent multi-modal causal mediation analysis for microbiome data have been implemented for SE in the multimedia package [@jiang2025], highlighting the readiness to support the adoption of the rapidly developing multi-omic methodology.

**Function- and sequence-level analyses.** Taxonomic analysis is frequently complemented by functional analyses, either based on functional predictions derived from amplicon or metagenomic sequences or from parallel metabolomic, metatranscriptomic or other functional assays. Our framework not only supports the incorporation of functional abundance tables within the same data objects, but many statistical methods are generic and can therefore be applied across different omics data types. Where appropriate, specialized methods are also available; for instance, MaAsLin3 [@nickols2026] and many other Bioconductor libraries support metatranscriptomics analysis, notame package [@klavus2020] for metabolome analyses recently added SE support, and the R for Mass Spectrometry (R4MS) project provides many interoperable tools. The framework has found recent applications in extended analyses of metagenomic features such as antimicrobial resistome composition [@salehi2025]. We also converted a large-scale compilation of resistome profiles in 8,972 metagenome samples [@lee2023] into the (Tree)SE format to support research and training in this area of microbiome analysis.

**Spatio-temporal, ecological, and epidemiological models.** Microbiome time
series [@berg2020] encompass a wide range of study designs, ranging
from short to long follow-up times, sparse to dense data, univariate
to multivariate and multi-omic monitoring
[@velten2022; @sherwani2025; @fukuyama2017],
and real, pseudo, or simulated time series. Moreover, an increasing number of
studies incorporate spatial dimension in microbiome analysis [@ruuskanen2021].
While spatio-temporal analyses require specialized methods, the miaTime
package offers a set of data wrangling methods to support longitudinal,
multi-modal microbiome analysis. In addition, Bioconductor contains applicable
methods from other research domains, such as single-cell omics
and spatial transcriptomics, which use closely
related data containers. Spatial and temporal information can be
incorporated as a covariate (*e.g.*, location or time) or by estimating
temporally or spatially resolved metrics, such as divergence (dissimilarity
between consecutive time points), autocorrelation (similarity of closely
related samples), and feature modality (whether a feature exhibits multiple
abundance states). Another type of temporal analysis
includes the time-to-event, or survival models, which are encountered
regularly in epidemiological and population cohort studies; for microbiome
data, these analyses require specialized approaches, as proposed in previous
studies [@salosensaari2021]. We also provide methods to simulate
time series from standard models of microbial community dynamics including
Hubbell's neutral model, generalized Lotka-Volterra, and the
consumer-resource model through the miaSim package [@gao2023]. Our framework
provides a robust basis for developing extended methods for longitudinal
multi-omics, which is currently an active area of investigation [@sherwani2025].
We also converted a large-scale time series of 16,234 gut microbiome profiles
from 585 wild baboons over 14 years [@grieneisen2021] into the
TreeSE format to support research and training on longitudinal datasets.

**Multi-modal integration.** A key feature of this framework is to provide the infrastructure supporting the integration of microbial abundance data with other omics modalities, such as transcriptomes, proteomes, or metabolomes [@hintikka2021; @mangnier2025], resistomes [@salehi2025; @lee2023], host genomes [@qin2022], or environmental exposomes (@fig-downstream_analyses). By leveraging standardized containers (Fig. \ref{fig:datacontainers}), users get fluent access to a growing number of multi-omic methods. Containers reduce the need for manual data wrangling and mitigate the risk of errors (*e.g.*, sample mismatches or data formatting), and support modular and reproducible workflows. They also facilitate incorporation of further downstream tools. Multi-omic methods are particularly relevant for the present ecosystem given its enhanced support for integrative analysis. Examples include association-based approaches, latent variable models, and Machine Learning (ML)-based prediction [@sankaran2019; @sankaran2019b; @sherwani2025]. Our framework is supported by many common multi-omic libraries in Bioconductor: Cross-correlation analysis, provided by mia, is commonly used to capture direct associations of samples or features between two datasets but it does not explicitly account for potential co-variation between microbial features, and methods such as anansi can additionally incorporate known functional relations into association analyses. Unsupervised latent variable methods such as multi-omics factor analysis (MOFA+) [@argelaguet2020] and joint robust PCA (Joint-RPCA) [@vargas2026] model shared latent structure across modalities, enabling the exploration and visualization of coordinated variation across multiple features and data sources. Supervised methods include for instance IntegratedLearner [@mallick2024], which can leverage dependencies between omics layers towards improved prediction performance. These and other methods natively support Bioconductor's multi-assay framework. Additional tools are provided by multi-omic toolkits such as the multi-omic discriminant analysis DIABLO [@singh2019] and mixOmics [@rohart2017], which do not yet support the same data containers but can be used seamlessly by using converters or a straightforward extraction of the abundance tables, sample metadata, or other relevant data elements. The unified integrative framework that we have adopted can accelerate methods development, benchmarking, and standardization.


**Other methods and programming environments.** Several other methods and workflows have been employed in microbiome data science. Our multi-assay framework provides a flexible statistical programming ecosystem dedicated to microbiome data science. It supports the incorporation of more targeted analysis methods and workflows such as SIAMCAT for association testing [@wirbel2021], mikropml for machine-learning based prediction and classification [@topcuoglu2021], NetCoMi for network analysis [@peschel2021], miaSim [@gao2023] for time series simulations, or microbe-set enrichment analyses [@kou2020; @pham2025]. These and other similar tools are complementary and they operate primarily on a single feature matrix, lacking the multi-omic feature selection offered by dedicated integration methods discussed in the previous section. Interoperable containers facilitate the translation of existing tools to adjacent or emerging disciplines.

Our framework can be linked to external microbiome data analysis pipelines using tools such as bioBakery [@beghini2021], and foreign platforms through language-specific interfaces, namely Rcpp for C++, reticulate for Python, and JuliaCall for Julia. The overall data analytical paradigm of Bioconductor is similar to that of scikit-bio [@aton2026], a Python-based ecosystem. Whereas scikit-bio benefits from the rapidly evolving machine learning and artificial intelligence tools in Python, Bioconductor's strengths lie in its vast statistical ecosystem and interoperability with dedicated omics tools and other computational environments. We anticipate that the ongoing developments in cross-language integration will further lower the barriers for cross-utilizing tools across complementary environments. This can be enhanced by additional workflow management tools such as *targets* in R [@landau2021] or *Nextflow* for cross-language operations [@di2017nextflow].

```{r}
#| label: downstream_analyses
#| echo: false
#| message: false
#| warning: false

# Configure Java for IntegratedLearner before loading the required packages
options(java.parameters = c("-Xmx6G", "--add-modules=jdk.incubator.vector"))


# Load required packages
pkgs <- c("IntegratedLearner", "maaslin3", "mia", "scater")

temp <- sapply(pkgs, function(x) {
    suppressMessages(require(x, character.only = TRUE))
})

# Helper for catching all printing from Maaslin3 and IL
quiet <- function(x) {
    invisible(capture.output(x))
    return(x)
}

# Load example multi-omics dataset
data("HintikkaXOData")
mae <- HintikkaXOData

# Attach sample metadata and perform basic preprocessing
mae[[1]] <- getWithColData(mae, 1)
mae[[2]] <- getWithColData(mae, 2)
assay(mae[[2]], "nmr")[is.na(assay(mae[[2]], "nmr"))] <- 0
rowData(mae[[1]]) <- apply(
    rowData(mae[[1]]),
    2,
    function(x) gsub("D_[1-6]__", "", x)
)

# Construct genus-level data and phylogenetic trees
mae[[1]] <- agglomerateByRanks(mae[[1]])
altExp(mae[[1]], "prev_genus") <- subsetByPrevalent(
    mae[[1]],
    assay.type = "counts",
    rank = "Genus",
    prevalence = 1,
    include.lowest = TRUE,
    detection = 1
)
mae[[1]] <- swapAltExp(mae[[1]], "Genus", "main")
mae[[1]] <- addHierarchyTree(mae[[1]])
altExp(mae[[1]], "prev_genus") <- addHierarchyTree(altExp(mae[[1]], "prev_genus"))
rowTree(altExp(mae[[1]], "prev_genus"))[["tip.label"]] <-
    gsub("Genus:", "", rowTree(altExp(mae[[1]], "prev_genus"))[["tip.label"]])

# Compute feature abundance summaries for tree visualization
altExp(mae[[1]], "prev_genus") <- addPerFeatureQC(
    altExp(mae[[1]], "prev_genus")
)
rowData(altExp(mae[[1]], "prev_genus"))[["log(mean abundance)"]] <-
    log(rowData(altExp(mae[[1]], "prev_genus"))[["mean"]])

# Generate transformed assays for downstream analyses
mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "counts",
    method = "relabundance",
    altexp = altExpNames(mae[[1]])
)
mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "counts",
    method = "rclr",
    altexp = altExpNames(mae[[1]])
)
mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "counts",
    method = "rclr",
    impute = FALSE,
    name = "rclr_non_imputed",
    altexp = altExpNames(mae[[1]])
)
mae[[2]] <- transformAssay(
    mae[[2]],
    assay.type = "nmr",
    method = "log10",
    pseudocount = TRUE
)

# Compute alpha diversity and ordination analyses
mae[[1]] <- addAlpha(mae[[1]], assay.type = "counts", index = "faith")
mae[[1]] <- addDissimilarity(
    mae[[1]],
    assay.type = "counts",
    method = "unifrac"
)
mae[[1]] <- addRDA(
    mae[[1]],
    dis.name = "unifrac",
    formula = x ~ Fat + XOS,
    na.action = na.exclude
)

# Perform differential abundance analysis with MaAsLin3
maaslin3_out <- maaslin3(
    input_data = mae[[1]],
    output = "maaslin_results",
    formula = "~ Fat + XOS"
) |> quiet()

maaslin3_res <- maaslin3_out[["fit_data_prevalence"]][["results"]]
maaslin3_res <- maaslin3_res[!is.na(maaslin3_res[["coef"]]), ]
maaslin3_res <- maaslin3_res[
    order(maaslin3_res[["qval_joint"]]),
]
maaslin3_res <- maaslin3_res[
    maaslin3_res[["qval_joint"]] < 0.05,
]

# Compute Joint-RPCA embedding across microbiome and metabolomics data
mae <- addJointRPCA(
    mae,
    experiments = c(1, 2),
    assay.types = c("rclr_non_imputed", "log10")
)

# Train an IntegratedLearner model for feature importance visualization
model <- IntegratedLearner(
    MAE_train = mae,
    experiment = c("microbiota", "metabolites"),
    assay.type = c("counts", "nmr"),
    outcome_col = "Fat",
    subject_id_col = "Sample",
    family = binomial()
) |> quiet()
```

```{r}
#| label: fig-downstream_analyses
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "**Selected single-omic and multi-omic visualizations for microbiome data.** The figure demonstrates visualization workflows available in the ecosystem using the openly available multi-omics mouse dataset distributed through the mia package [@hintikka2021]. The dataset contains microbiota (16S rRNA) and metabolomics measurements from mice receiving either a high-fat or low-fat diet, with or without xylooligosaccharide (XOS) supplementation. **a**, Taxonomic relationships among bacterial genera represented as a tree. Most genera belong to the phylum *Firmicutes*, while *Alistipes* and *Parabacteroides* have the highest mean abundances. **Single-omic analyses (b–d)** demonstrate common analyses of individual microbiome data layers. **b**, Faith's phylogenetic diversity [@armstrong2021] summarizes within-sample microbial diversity and shows no clear difference between diet groups. **c**, Distance-based redundancy analysis (dbRDA) based on UniFrac distances [@lozupone2005] visualizes differences in microbial community composition explained by the experimental variables, revealing separation according to diet. **d**, Differential prevalence analysis performed with MaAsLin3 [@nickols2026] identifies bacterial genera whose prevalence differs significantly between diet groups (adjusted *P* < 0.05). **Multi-omic analyses (e–f)** integrate microbiota and metabolomics data. **e**, Joint-RPCA provides an unsupervised low-dimensional representation of the combined microbiota and metabolomics data. The five strongest loading vectors from each data layer indicate the features contributing most to the displayed ordination axes. Vectors pointing in similar directions indicate positively associated features, whereas vectors pointing in opposite directions indicate negative associations. **f**, Signed feature importances from an IntegratedLearner model quantify the contribution of individual microbiota and metabolite features to supervised prediction of dietary group, indicating that metabolite features are generally more predictive than microbiota features, with glucose identified as the most important predictor."
#| fig-width: 20
#| fig-height: 15

pkgs <- c("dplyr", "ggtree", "grid", "miaViz", "patchwork")

temp <- sapply(pkgs, function(x) {
    suppressMessages(require(x, character.only = TRUE))
})

pkgs <- c("dplyr", "ggtree", "miaViz", "patchwork")

temp <- sapply(pkgs, function(x) {
    suppressMessages(require(x, character.only = TRUE))
})

# Create individual panels -----------------------------------------------------

# Phylogenetic tree coloured by phylum and mean abundance
p_tree <- plotRowTree(
    altExp(mae[[1]], "prev_genus"),
    edge.colour.by = "Phylum",
    show.label = rowTree(altExp(mae[[1]], "prev_genus"))$tip.label,
    tip.colour.by = "log(mean abundance)",
    label.font.size = 2
)

# Alpha diversity
p_alpha <- plotBoxplot(
    mae[[1]],
    col.var = "faith", x = "Fat", colour.by = "Fat"
)

# Distance-based redundancy analysis
p_rda <- plotRDA(
    mae[[1]], "RDA",
    colour.by = "Fat", fill.by = "Fat",
    add.ellipse = TRUE,
    add.significance = FALSE,
    add.expl.var = TRUE
)

# Differential abundance results
p_forest <- plotForest(
    maaslin3_res,
    effect.var = "coef",
    err.var = "stderr",
    pval.var = "qval_joint",
    id.var = "feature",
    label.by = c("CI", "P-Value"),
    order.by = "coef"
)

# Joint-RPCA ordination
p_joint_rpca <- plotJointRPCA(
    mae, "JointRPCA",
    colour.by = "Fat",
    fill.by = "Fat",
    add.ellipse = TRUE,
    ntop = 5,
    add.expl.var = TRUE,
    label.size = 2
)

# Prepare feature importance data ----------------------------------------------

df <- bind_rows(
    lapply(names(model[["feature_importance_signed_by_layer"]]), function(layer) {
        x <- model[["feature_importance_signed_by_layer"]][[layer]]
        data.frame(
            Layer = layer,
            Feature = names(x),
            Effect = unname(x)
        )
    })
)

# Keep the strongest positive and negative feature associations
df <- df |>
    mutate(
        Magnitude = abs(Effect),
        Direction = ifelse(Effect >= 0, "Positive", "Negative")
    ) |>
    group_by(Layer) |>
    slice_max(Magnitude, n = 20, with_ties = FALSE) |>
    arrange(Layer, desc(Magnitude)) |>
    mutate(Feature = factor(Feature, levels = Feature)) |>
    ungroup()

# Feature importance lollipop plot
p_feature_importances <-
    ggplot(df, aes(Magnitude, Feature, colour = Direction)) +
    geom_segment(
        aes(x = 0, xend = Magnitude, yend = Feature),
        linewidth = 0.8
    ) +
    geom_point(size = 2) +
    facet_wrap(~Layer, scales = "free") +
    scale_colour_manual(values = c(
        Positive = "firebrick3",
        Negative = "steelblue3"
    )) +
    scale_x_log10() +
    labs(
        x = "Feature importance",
        y = NULL,
        colour = "Association"
    ) +
    theme_bw()

# Harmonize the appearance of all panels ---------------------------------------

common_theme <- theme(
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    strip.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 11),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5)
)

p_alpha <- p_alpha + common_theme
p_forest <- p_forest + common_theme
p_rda <- p_rda + common_theme

# Place the linetype legend inside the Joint-RPCA panel
p_joint_rpca <- p_joint_rpca +
    common_theme +
    theme(
        legend.position = "inside",
        legend.position.inside = c(0.98, 0.02),
        legend.justification = c(1, 0),
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.key = element_blank()
    )

p_tree <- p_tree + common_theme
p_feature_importances <- p_feature_importances + common_theme

# Fine-tune individual panels --------------------------------------------------

# Reduce tree width and move legends inside the panel
p_tree <- p_tree +
    xlim_tree(1) +
    guides(
        fill = guide_colorbar(
            position = "inside",
            title.position = "top",
            theme = theme(
                legend.position.inside = c(1.5, 0.5),
                legend.justification = c(0, 1)
            ),
            barheight = unit(2.5, "cm"),
            barwidth = unit(0.4, "cm")
        ),
        colour = guide_legend(
            position = "inside",
            title.position = "top",
            theme = theme(
                legend.position.inside = c(1.5, 0.5),
                legend.justification = c(0, 1)
            )
        )
    )

# Remove redundant legends and harmonize axis labels
p_alpha <- p_alpha +
    labs(y = "Faith", x = "Fat") +
    guides(colour = "none", fill = "none")

p_rda <- p_rda +
    guides(colour = "none", fill = "none") +
    theme(aspect.ratio = 1 / 2)

p_joint_rpca <- p_joint_rpca +
    guides(colour = "none", fill = "none") +
    theme(aspect.ratio = 1 / 2)

p_forest <- p_forest[[1]]
p_forest <- p_forest +
    labs(
        x = "Association with high-fat diet (log-odds)",
        y = ""
    )

# Add panel labels -------------------------------------------------------------
# We have to add tags manually because when using patchwork::plot_annotation(),
# they would be added also to titles.

tag_theme <- theme(
    plot.tag = element_text(face = "bold", size = 15),
    plot.tag.position = c(0.01, 0.99)
)

p_tree <- p_tree + labs(tag = "a") + tag_theme
p_alpha <- p_alpha + labs(tag = "b") + tag_theme
p_rda <- p_rda + labs(tag = "c") + tag_theme
p_forest <- p_forest + labs(tag = "d") + tag_theme
p_joint_rpca <- p_joint_rpca + labs(tag = "e") + tag_theme
p_feature_importances <- p_feature_importances + labs(tag = "f") + tag_theme

# Assemble the figure ----------------------------------------------------------

panel_single_omic <-
    (p_alpha + p_rda + p_forest +
        plot_layout(widths = c(1, 1, 1))) +
    plot_annotation(title = "Single-omics")

panel_multi_omics <-
    (p_joint_rpca + p_feature_importances +
        plot_layout(widths = c(1, 2))) +
    plot_annotation(title = "Multi-omics")

# Standalone section headers
title_single <- wrap_elements(
    full = textGrob(
        "Single-omics",
        x = 0,
        hjust = 0,
        gp = gpar(fontface = "bold", fontsize = 13)
    )
)

title_multi <- wrap_elements(
    full = textGrob(
        "Multi-omics",
        x = 0,
        hjust = 0,
        gp = gpar(fontface = "bold", fontsize = 13)
    )
)

# Combine all panels into the final figure
final <-
    p_tree /
    title_single /
    panel_single_omic /
    title_multi /
    panel_multi_omics +
    plot_layout(heights = c(3, 0.08, 1, 0.08, 1))

final
```

# Accessible analysis and community {#sec-community}

```{r}
#| label: bioc_views2
#| eval: true
#| echo: false
bioc_pkgs <- read.csv(file.path("data", "packages.csv"))
```

Modern science is embodied by a critical community capable of
assessing discoveries and replicating results [@wootton2016].
Open source developer communities present a timely manifestation of this
principle, with technical, legal, sociological and epistemic consequences
[@hocquet2024]. FAIR principles for software support inclusive
access to well-documented and user-friendly methods [@barker2022],
and are facilitated by the extensive quality control and
documentation standards of Bioconductor [@huber2015].

In addition to these technical demands, the development and quality control of research software rely significantly on community engagement. This stimulates continuous software improvement and promotes the emergence and dissemination of evidence-based best practices [@berg2020;
@barker2022; @hocquet2024]. Simple approaches should be preferred over complex ones by default, and key analyses should be complemented with visual exploration and verification. Expressive code and meaningful comments support reproducibility and verification. Such practices can greatly improve the transparency and robustness of the analyses and interpretability of the results.
OMA facilitates standardization across studies by promoting the use of harmonized data structures, validated methods, and informed default settings. This facilitates methods sharing and benchmarking, helping to establish community best practices and build consensus while allowing users to flexibly select and evaluate methods that are appropriate for their specific data and research questions. The Bioconductor framework naturally aligns with these goals through its open statistical programming ecosystem that encompasses data resources, interoperable software, and educational materials. Community building is actively supported through online training resources and communication channels, making it easier for both developers and users to contribute to the continued critical assessment, development and standardization of the methods and documentation. Together with interactive applications, these resources offer accessible entry points for developers and users regardless of their programming experience.

**The OMA book.** _Orchestrating Microbiome Analysis with Bioconductor_ (OMA)
is an online book published at [https://bioconductor.org/books/release/OMA](https://bioconductor.org/books/release/OMA) that provides a comprehensive guide to the (Tree)SE framework for microbiome data science in Bioconductor, including practical guidelines,
reproducible workflows, and training material for integrative microbiome
analyses. The book is versioned according to the Bioconductor's biannual
release cycle. The work incorporates rich feedback from the user community and
microbiome training courses. The OMA book aligns to a broader Bioconductor convention, where prominent data structures [@huber2015; @ramos2017; @huang2021]
designed for multi-table data integration have been described across several
domains, such as single-cell omics (OSCA book [@amezquita2020]) and spatial transcriptomics (OSTA book). These resources use standard data containers to support statistical programming on interlinked multi-table datasets.

**Interactive applications.** Graphical user interfaces provide accessible entry points. The iSEEtree and miaDash packages provide interactive analysis and visualization tools for microbiome data [@benedetti2025]. The former extends iSEE to provide support for
TreeSE objects, while the latter builds on iSEEtree to enable analyses without requiring any programming. Their interface supports automatic generation of source code and replicable workflows and analysis templates for non-technical users. Moreover, the tidy R paradigm, available through tidyomics, simplifies the syntax and streamlines analytical workflows [@hutchison2024]. These tools allow users to analyze and visualize complex data combinations with ease.

**Community.** Its widespread adoption has made R one of the most cited software resources of our time [@pearson2025]. The Bioconductor community is widely recognized for developing open research software for life sciences. Its strengths include minimal documentation standards, portability across operating platforms, quality control (*e.g.* peer-review, unit testing, and continuous integration), usability (APIs), robustness (dependency testing and deprecation mechanisms), and a biannual release cycle with semantic versioning. Over the past two decades Bioconductor has grown into a repository of more than 2,300 actively maintained software packages, including
`r sum(bioc_pkgs$Field == "Microbiome")` packages related to microbiome research. At the core of the Bioconductor's data science strategy has been the idea of shared data representations that can enhance the transparency, trustworthiness, and explainability of analyses [@gentleman2004; @lawrence2013]. This has mitigated redundant efforts and facilitated the translation of methods between research domains [@delia2023; @barker2022]. The standards have also played an important role in the formation of broad developer and user communities [@drnevich2025]. Bioconductor has a strong tradition in fostering the community through shared data structures and training activities [@drnevich2025]. An active community of developers and users thrives online, supporting the software sustainability and promoting good practices in scientific computing [@wilson2017]. The community interacts via several online communication channels, including Bioconductor Zulip, GitHub platform, and an email list. The development is influenced by feedback from the user community through these channels, regular meetings, and training workshops.


# Outlook {#sec-discussion}

```{r}
#| label: bioc_views
#| eval: false
#| echo: false
#| message: false

if (!require("BiocPkgTools")) {
    BiocManager::install("BiocPkgTools")
    library("BiocPkgTools")
}
# Get packages and info based on their BiocViews
bioc_pkgs <- biocPkgList()
#
fields <- sapply(bioc_pkgs[["biocViews"]], function(x) {
    field <- "Other"
    if (any(grepl("genomic|genetic|gene|genom|DNA", x, ignore.case = TRUE))) {
        field <- "Genomics"
    }
    if (any(grepl("proteomics|protein", x, ignore.case = TRUE))) {
        field <- "Proteomics"
    }
    if (any(grepl("metabolomics|metabolome|metabolite|Lipidom|massspectro", x, ignore.case = TRUE))) {
        field <- "Metabolomics"
    }
    if (any(grepl("transcripto|RNA-seq|RNA", x, ignore.case = TRUE))) {
        field <- "Transcriptomics"
    }
    if (any(grepl("immuno", x, ignore.case = TRUE))) {
        field <- "Immunology"
    }
    if (any(grepl("cytom", x, ignore.case = TRUE))) {
        field <- "Cytometrics"
    }
    if (any(grepl("microarray|chip", x, ignore.case = TRUE))) {
        field <- "Microarray"
    }
    if (any(grepl("single-cell|singlecell", x, ignore.case = TRUE))) {
        field <- "Single-cell"
    }
    if (any(grepl("metagenom|microbiome|16S|microbiota|amplicon|shotgun|microb|metatranscript|metametabolo|metaproteo", x, ignore.case = TRUE))) {
        field <- "Microbiome"
    }
    return(field)
})
bioc_pkgs[["Field"]] <- fields

bioc_pkgs[, c("Package", "Field")] |>
    write.csv(file = file.path("data", "packages.csv"), row.names = FALSE)
```

**Microbiome research** has shifted towards integrative analyses
linking various omics towards enhanced mechanistic and causal insights [@joos2024; @bastiaanssen_bugs1_2023]. Technological advances in long-read sequencing, functional assays, spatial profiling, and single-cell analysis have created an increasing demand for integrative techniques that link microbiome data with other omics along varying spatial and temporal dimensions. Future extensions could thus provide guidance for longitudinal and repeated-measures microbiome studies, where temporal dependency and intra-subject correlation introduce additional analytical complexity. The field could also benefit from improved support for the FAIR data principles [@wilkinson2016], for instance by adding built-in refinement, enrichment, and validation methods for community standards [@field2008; @mirzayi2021; @kelliher2025].

Moreover, there is a need for integrated multi-omics workflows that can combine microbiome data with other omics to support interpretable hypotheses and causal analyses. In concert with parallel developments in other omics, our framework specifically addresses the need to jointly analyze interlinked abundance tables while supporting hierarchical and tabular side information. Future developments may necessitate the adoption of extended hierarchical data structures such as Bioconductor's GraphExperiment.

Our framework provides the structured input necessary for probabilistic inference and scalable computation as the recent developments increasingly rely on advanced statistical and ML techniques such as Bayesian modeling and Gaussian processes to accommodate uncertainty and temporal or spatial structures [@jeganathan2021; @moreno-indias2021].

Building upon these foundations, the rapid emergence of foundation models and deep learning approaches in biomedicine calls for robust, scalable data representations such as those we have adopted. Applications include transfer learning across studies,
interpretable embeddings of microbiome multi-omic profiles,
computational prediction of microbial function and metabolites, and integration with AI agents for hypothesis generation and automated analysis, among others [@delia2023]. The interoperability between R and other programming languages could further facilitate the application of novel ML tools and frameworks.


**Open data science infrastructures** have become critical components of reproducible research and microbiome applications critically rely on open-source methods [@berg2020; @barker2022;
@hocquet2024]. To enjoy the benefits of automation while also maintaining flexibility, data scientists often recombine open-source software components into custom workflows. However, the lack of commonly adopted data representation standards can hinder the development, sharing, and application of modular workflows.

Building upon a coordinated community effort, we have reported the development of a comprehensive methods ecosystem for multi-modal microbiome data science in R/Bioconductor, built on widely adopted Bioconductor data containers, software libraries, open data resources, tutorials, and a growing community of developers.

The users can benefit from an extended collection of end-to-end case studies demonstrating how different example analyses can be conducted in practice, from original source data to results. Hence, we have provided comprehensive guidance in the OMA online book, to support independent learning and reproducible analysis in microbiome research and beyond [@timmis2024; @drnevich2025]. We welcome new community contributions of well-documented methods and workflow demonstrations. As the field evolves, the developer communities will continue to integrate new tools and approaches to improve integrative analyses of microbiome data.



## Data availability

All datasets used in this article are publicly available with a permanent digital object identifier and an open source license from the specified sources.


## Code availability

This article presented v1.1.1 of the OMA book (2026-07-05) published at
[https://bioconductor.org/books/release/OMA](https://bioconductor.org/books/release/OMA),
which depends on R v4.6 and Bioconductor v3.24. Analysis scripts and
results of the benchmarking analysis are available on Zenodo (mia community).

# References
<!-- the pattern below controls the placement of the references -->
:::{#refs}
:::

## Consortia {#sec-consortium}

OMA consortium: Arora, Nalin; Bastiaanssen, Thomaz; Beber, Moritz E.;
Bektanov, Aituar; Benedetti, Giulio; Bhowmick, Sabuj; Borman, Tuomas;
Braccia, Domenick J.; Calgaro, Matteo; Callan, Danielle; Chaudhary,
Jiya; Cordazzo Vargas, Bianca; Corrada Bravo, Héctor; Courbayre
Dussau, Basil; Das, Sneha; Eckermann, Henrik; Erawijantari, Pande;
Ernst, Felix G.M.; Gaudron-Parry, Elliot; Gopalakrishnan, Shyam;
Hakula, Maija; Hanhineva, Kati; Havulinna, Aki; Hindström, Rasmus;
Jeba, Akewak; Lahti, Leo; Limborg, Morten; Lindgren, Himmi; Liu,
Yihan; Mallick, Himel; Martino, Cameron; Mongad, Dattatray; Müller,
Christian L.; Muluh, Geraldson; Shenhav, Liat; Nynäs, Signe; Pasanen,
Jesse; Paulo Casucci dos Santos, Joao; Pelto, Juho; Peschel, Stefanie;
Potbhare, Renuka; Pralas, Théotime; Ramos, Marcel; Raulo, Aura;
Sannikov, Artur; Seraidarian, Ely; Serizay, Jacques; Shetty,
Sudarshan; Shigdel, Rajesh; Suksi, Vilhelm; Tammi, Eetu; Valderrama,
Benjamin; Waldron, Levi; Yang, Lu.



## Author contributions {#sec-contributions}

L.L. and T.Bo. conceived and coordinated the study. L.L., T.Bo., G.B., A.R., A.S., R.H. and T.Ba. performed the data analysis. L.L., T.Bo. and G.B. wrote the original draft. L.L. acquired funding. All authors contributed material, participated in manuscript preparation and approved the final version.


## Competing interests {#sec-interests}

The authors declare no competing interests.

## Acknowledgments {#sec-acknowledgments}

We thank CSC – IT Center for Science, Finland, for computational resources.

This project received funding from the European Union’s Horizon 2020
research and innovation programme under grant agreement No 952914 and
Research Council of Finland (decision 374944). TB was additionally
supported by the University of Turku Graduate School.
We thank
Christopher Quince and Kihyun Lee for providing resistome profiling data,
Justin Silverman for PhILR support,
Ben Allen and Sandy Rogers for MGnifyR support,
Elizabeth Archie and Johannes Björk for support with the wild baboon data set,
and the following individuals for supporting the project:
Richa Ashma; Nitin Bayal; Chouaib Benchraka; Karoline Faust; Robert D. Finn; Samuel Hillman; Ruizhu
Huang; Lorna Richardson; Daniel Rios Garza; Matti Ruuskanen; Mahkameh Salehi.
We extend our thanks to all contributors of the open-source resources used in this work.
